---
title: "AAC Import/clean"
output:
  html_document:
    df_print: paged
---

## Goals

The goals of this notebook are:

- Are animal intakes increasing or decreasing each year since 2016? Plot a column chart of intakes by year. Don’t include 2021 since it is not a full year.
- Are there seasonal monthly trends in overall animal intakes? Plot intakes by year/month from 2016 to current, including 2021. (One long line or column chart.)
- Plot that same monthly trend, except with month as the x axis, with a new line for each year.
- Do certain types of animals drive seasonal trends? Use five full years of data (not 2021) to summarize intakes by animal type and month. Plot with month on the x axis with a line for each animal type.


```{r setup, setup, setup, echo=T, results='hide', message=F, warning=F}
library(tidyverse)
library(janitor)
library(lubridate)
```

# Import data

Import data downloaded from [austintexas.gov website](https://data.austintexas.gov/)

```{r import}
raw_data <- read_csv("data-raw/Austin_Animal_Center_Intakes.csv") %>% clean_names() #import csv, clean up column names

# Peek at data
raw_data %>% glimpse()
```

## Fix dates

Clean up the dates in the dataset so they're usable for analysis

```{r fix_dates}
date_fix <- raw_data %>% 
  mutate(
    intake_date = str_sub(date_time, 1, 10), # Grab substring with date only
    intake_date = mdy(intake_date) # convert to date
  )

date_fix %>% glimpse()
``` 

# New date helper columns

Use lubridate functions to get new `yr`, `mo`, and `yrdate` columns from the `intake_date` to help with our analysis later

```{r date_chunks}
date_parts <- date_fix %>% 
  mutate(
    yr = year(intake_date), # create month column
    mo = month(intake_date, label = TRUE), # create month column
    yr_mo = format(intake_date, "%Y-%m") # Extract year plus month as a column
  ) %>% 
  select(
    intake_date, yr, mo, yr_mo, everything(), -date_time, -month_year
  )

date_parts %>% glimpse()
```

## Filter for dates

Filter the data to only animals intakes since 2016 

```{r filter_dates}
filtered_dates <- date_parts %>% 
  filter(
    intake_date > "2015-12-31", # filter for intakes from 2016 onward
    intake_date < "2021-10-01" # exclude incomplete current month
  )

# check to make sure dates filtered correctly
filtered_dates %>% summary()
```

## Export 

Export our cleaned data so we can reference it in our separate analysis notebook 

```{r export}
filtered_dates %>% 
  write_rds("data-processed/01-intakes.rds")
```