---
title: "Austin Animal Center Data"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

## About the project

This project analyzes animal intake data from Austin Animal Center since 2016. The data was imported from [Austin's Open Data Portal](https://data.austintexas.gov/), and cleaned separately by me.

## Setup

Import the following libraries to help analyze our data: `tidyverse`, `lubridate`, `scales`, `plotly`, and `clipr`

```{r setup, setup, echo=T, results='hide', message=F, warning=F}
library(tidyverse)
library(lubridate)
library(scales)
library(plotly)
library(clipr)
```

## Import data

Import the Austin Animal Center intakes dataset we cleaned in `01-import`

```{r import}
intakes <- read_rds("data-processed/01-intakes.rds")

# peek at the data
intakes %>% glimpse
```

## Intakes by year

Answer the following question: 

- Are animal intakes increasing or decreasing each year since 2016? Plot a column chart of intakes by year. Don’t include 2021 since it is not a full year.

### Summarise data

```{r summarise_intakes_by_yr}
intakes_yr_data <- intakes %>% 
  filter(
    yr < "2021"
  ) %>% 
  count(yr, name= "count_animals")

intakes_yr_data 
```
### Plot data

```{r plot_intakes_by_yr}
ggplot(intakes_yr_data, aes(x = yr, y = count_animals)) + 
  geom_col(aes(fill = yr)) +
  geom_text(aes(label = comma(count_animals)), vjust = 2, color = "white") +
  labs(
    title = "Intakes By Year",
    subtitle = "Total animals taken in by Austin Animal Center per year since 2016",
    caption="Source: austintexas.gov",
    x = "Year",
    y = "Total animals"
  ) +
  scale_y_continuous(labels = comma)
```

## Intakes by month

Answer the following question:

- Are there seasonal monthly trends in overall animal intakes? 

### Summarise the data

Use `count` to find monthly intakes by year

```{r summarise_intakes_by_month}
intakes_yrmo <- intakes %>% 
  count(yr_mo, name = "count_animals")

intakes_yrmo
```

### Plot intakes by yr/mo

Look at the overall trends of intakes broken down by month to understand where we should dig in more.

```{r plot_intakes_by_month, fig.width = 10}
ggplot(intakes_yrmo, aes(x = yr_mo, y = count_animals)) +
  geom_line(group = 1) +# tell ggplot to only plot the single line
  theme(axis.text.x = element_text(angle = 90)) # flips the axis name
```

## Intakes by month, split by year

Use month and year columns to show trends in intakes by year

### Summarise the data

Use `count` to intakes by month and year intakes by year

```{r count_mo_yr}
mo_yr_data <- intakes %>% 
  count(mo, yr, name = "count_animals") # count animal intakes by month, year

mo_yr_data
```

### Plot

```{r plot_mo_yr_data}
ggplot(mo_yr_data, aes(x = mo, y = count_animals)) + # set x and y axis
  geom_line(aes(group = yr, color = as.character(yr))) + # split lines and set colors
  labs(
    title = "Intake Trends By Month",
    subtitle = "Amount of animals taken in monthly by year",
    x = "Month",
    y = "Animal Count",
    color = "Year", # fix the title for the legend
    caption = "Source: austintexas.gov"
  )
```

## Types by month

Answer the following question (using only full-year data)

- Do certain types of animals drive seasonal trends? 

### Recode data

Create an "other" group for animals other than cats and dogs, filter data for complete years

```{r recode_types}
intake_types <- intakes %>% 
  filter(yr < 2021) %>% 
  mutate(
    animal_type = recode(
      animal_type, # column
      "Bird" = "Other", # Values to rename
      "Livestock" = "Other"
      )
  )

intake_types %>% 
  count(animal_type)
```

### Summarise data

Use `count` to find amount of animals taken in by month

```{r types_by_month}
type_mo_count <- intake_types %>% 
  count(mo, animal_type, name = "count_animals") # count animal types per month

# peek at the data
type_mo_count
```

### Plot animal types by month

```{r plot_type_per_mo}
ggplot(type_mo_count, aes(x = mo, y = count_animals)) +
geom_line(aes(group = animal_type, color = animal_type)) + # split by animal type
  labs(
    title = "Monthly Trends In Animal Types",
    subtitle = "Types of animals taken in by type each month since 2016",
    x = "Month",
    y = "Animal Count",
    color = "Type", # fix the title for the legend
    caption = "Source: austintexas.gov"
  )
```

